Effective Reionization Likelihood from Planck 2018 Data (?)

### 2021-03-22 paper/rei.tex

The cosmic microwave background (CMB) has entered an era of precision cosmology. The measurements from the \textit{Planck} satellite have shown agreement with the standard $\Lambda$CDM model which describes the initial perturbations in the Universe and their evolution. While many components of the standard cosmology model is well-understood, the details of the process of reionization however, remains one of the most uncertain pieces. Its uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude. Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB; and will also have implications for inferring cosmic acceleration through the growth of structure. [add more refs here]

@@

In fact, to extract the most information possible from this E-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum. This method has been applied to WMAP and Planck data to obtain complete constraints on reionization models~\cite{Mortonson:2008rx, Mortonson:2007hq, Heinrich:2016ojb, Aghanim:2018eyx}. It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, as well as massive neutrinos and gravitational waves in Ref~\cite{Dai:2015dwa}[also cite our inflation papers?].

@@

Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood, called $\texttt{srollv2}$, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the steplike model is reduced by roughly a factor of two from XX to XX. [fill in numbers]

@@

The code is publicly available on GitHub\footnote{[give link]}. [a bit more description here]. [comment on speed]. [comment on joint likelihood].

Extending PCs to cover up to $z_{\rm max} = 50$,we verified that there is no hint of ionization beyond $z>30$ in this Planck data release. But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is XX times less stringent than the Planck official results using Flex Knots [fill in number and make sure to say its flex knot or PC] in Ref.~\cite{Aghanim:2018eyx}. Moreover, we demonstrate that the PC results of Ref.~\cite{Aghanim:2018eyx} suffers from a logical flaw in the treatment of its priors following Ref.
~\cite{Millea:2018bko} and ends up penalizing lower optical depth models. [quote difference in tau(15, 30) or total tau possibly attributed to this treatment]. %We follow the appendix of Ref.~\cite{Heinrich:2018btc} for the Planck 2015 data

The paper is structured as follows. We first describe in section~\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood. Then in section~\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and compare with previous results in literature. Then in section~\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes. Finally, we summarize our results and conclude in section~\ref{sec:conclusion}.

@@

\section{PC Technique}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. We now briefly summarize the PC technique in section~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in section~\ref{sec:KDE}. We refer the readers to Refs.~\cite{Hu:2003gh, Mortonson:2008rx, Mortonson:2007hq, Heinrich:2018btc}[pick 1-2] and Refs.~\cite{Heinrich:2016ojb} respectively for a more complete description on these topics.

%
%
 \begin{figure}
          \includegraphics[width=0.48\textwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
          \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$. We assume fully ionized Universe at $z < 6$. \textit{Bottom}: The cumulative optical depth resulting from a unit amplitude PC. \ch{plot done}} 
          \label{fig:xe}
\end{figure}

\subsection{Reionization Principal Components}
\label{sec:PC}
We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
%
\begin{equation}
x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
\label{eq:mmutoxe}
\end{equation}
%
where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model. We obtain the PC basis functions $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements 
%
\beq
F_{ij} = \sum_l \frac{2 l+1}{2} \frac{\partial \ln C_l^{EE}}{\partial x_e(z_i)}\frac{\partial \ln C_l^{EE}}{\partial x_e(z_j)} = \sum_a \frac{ S_a(z_i) S_a(z_j)}{\sigma_a^2},
\eeq
%
where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs.
The Fisher matrix is computed at a fiducial model which we take to be $x_e^{\rm fid}(z)=0.15$ in the parameterized range.
 Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the PCs to be computed with Planck 2015 rather than WMAP best-fit 5 remaining $\Lambda$CDM parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here.

We then rank-ordered the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ for the fiducial analysis.\footnote{We also test the robustness of our fiducial analysis in \S XX by  taking 
 for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.}

 In Fig.~\ref{fig:xe}
 %
 \footnote{There are very small differences between the PCs here and those shown in Ref.~\cite{Heinrich:2016ojb}, which can be attributed to a chance in the best-fit cosmology (WMAP to Planck 2015) when constructing the PCs.}
 %
 (top), we show the corresponding fiducial basis functions $S_a(z)$.  
Notice that the series resembles a Fourier decomposition of $x_e(z)$ rank ordered from low to high frequency.
This is because high frequency variations in $x_e$ leave little observable imprint in $C_l^{EE}$.  
As a consequence  the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
For the Planck data set, most of the information on the ionization history comes from contraints on the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}. 
For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a transition width $\Delta z = 0.5$ (see also \S \ref{sec:example1}).

@@

@@

\subsection{Effective PC Likelihood}
\label{sec:KDE}
The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data. In the following, we briefly recap the kernel density estimate (KDE) technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~(\ref{eq:xetommu}). Since $\mathbf{m}$ can take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
% 1.06 cut w/ physicality prior
% 1.072 no cut
For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
%Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors. 
Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p).
\end{equation}

This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
For comparison, an even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
%
\begin{equation}
 {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
 \label{eq:gaussian}
 \end{equation}
%
In section~\ref{sec:effective_likelihood} we test the effective KDE and Gaussian approaches against a direct implementation
for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

@@

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{align}
\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)}.
\end{align}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.   \wh{Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
\begin{align}
\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^5 m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
\label{eq:cumtauPC}
\end{align}
where
\begin{align}
    \tau_a(z,z_{\rm max}) &= 
    n_{\rm H}^{\rm fid}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{S_a(z) (1+z)^{2} }{H_{\rm fid}(z)}, \nonumber\\
    \tau_{\rm fid}(z,z_{\rm max}) &= 
    n_{\rm H}^{\rm fid}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e^{\rm fid}(z) (1+z)^{2} }{H_{\rm fid}(z)},
\end{equation}
 where ``fid" refers to evaluation in the fiducial cosmology.   When we omit the redshift arguments, we implicitly mean the total range, e.g. $\tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max})$. }

 \wh{
 In Fig.~\ref{fig:xe}, we display $\tau_a(z,z_{\rm max})$.
 Notice that positive values of the first component $m_1$ mainly provides the optical depth that is accumulated at high redshift whereas {\it negative} values of $m_2$ provides the same for low redshift.    Since these two are the best measured components, the CMB mainly determines the amount of low vs.~high redshift ionization.
The underlying reason is that in $C_l^{EE}$ the higher the redshift, the
larger the relative contribution at higher $l$ due to the size of the horizon when the photons scattered.  For the less well measured PCs with $a>2$, the contributions to the total optical depth rapidly diminish and represent finer distinctions in redshift for the cumulative optical depth.}

\wh{
In the usual approach to constraining reionization in $\Lambda$CDM, one places an implicit prior on the shape of the cumulative optical depth by choosing a model like a near step function reionization and then extracts a single constraint on the total optical depth.}
%For example, the standard steplike model adopted in CAMB uses a width=0.48 function to parameterize the hydrogen and singly ionized helium reionization
% \begin{equation}
%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \width=0.48\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
% \label{eqn:tanh}
% \end{equation}
% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
% excludes finite ionization above its transition redshift by definition.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
unphysical ionization fractions where $0 \le x_e \le x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a  prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\label{eq:prior_sum}
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
\wh{Here we take $x_e^{\rm max} = 1+ n_{\rm He}/n_{\rm H}$ to account for the contribution of singly ionized helium.}
In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
% I think it can't depend on the data but does depend on the truncation
%the Planck 2018 data \wh{and
our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.

%
Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for a physical ionization history.  
In other words, no physical model is excluded by these priors but some unphysical models are included.  In \S \ref{sec:note_on_priors} we further explore the role of these and other choices of PC priors, and in \S\ref{sec:effective_likelihood} we test PC constraints against direct constraints for two example models.

@@

>>>>

The cosmic microwave background (CMB) has entered an era of precision cosmology. The measurements from the \textit{Planck} satellite have shown agreement with the standard $\Lambda$CDM model which describes the initial perturbations in the Universe and their evolution. While many components of the standard cosmology model is well-understood, the details of the process of reionization however, remains one of the most uncertain pieces. Its uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude. Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB; and will also have implications for inferring cosmic acceleration through the growth of structure. \ch{[add more refs here]}

@@

In fact, to extract the most information possible from this E-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum. This method has been applied to WMAP and Planck data to obtain complete constraints on reionization models~\cite{Mortonson:2008rx, Mortonson:2007hq, Heinrich:2016ojb, Aghanim:2018eyx}. It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, as well as massive neutrinos and gravitational waves in Ref~\cite{Dai:2015dwa}\ch{[also cite our inflation papers?]}.

@@

Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood, called $\texttt{srollv2}$, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the steplike model is reduced by roughly a factor of two from \ch{XX to XX. [fill in numbers]}

@@

\ch{write this part}
The code is publicly available on GitHub\footnote{[give link]}. [a bit more description here]. [comment on speed]. [comment on joint likelihood].

Extending PCs to cover up to $z_{\rm max} = 50$,we verified that there is no hint of ionization beyond $z>30$ in this Planck data release. But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is XX times less stringent than the Planck official results using Flex Knots \ch{[fill in number]} in Ref.~\cite{Aghanim:2018eyx}. Moreover, we demonstrate that the PC results of Ref.~\cite{Aghanim:2018eyx} suffers from a logical flaw in the treatment of its priors following Ref.
~\cite{Millea:2018bko} and ends up penalizing lower optical depth models. [quote difference in tau(15, 30) or total tau possibly attributed to this treatment]. %We follow the appendix of Ref.~\cite{Heinrich:2018btc} for the Planck 2015 data

The paper is structured as follows. We first describe in \S~\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood. Then in \S~\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and compare with previous results in literature. Then in \S~\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes. Finally, we summarize our results and conclude in \S~\ref{sec:conclusion}.

@@

\section{PC Technique}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. We now briefly summarize the PC technique in \S~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in \S~\ref{sec:KDE}. We refer the readers to Refs.~\cite{
%Hu:2003gh, %Mortonson:2008rx, 
Mortonson:2007hq,Heinrich:2016ojb} respectively for a more complete description of these topics.

\subsection{Reionization Principal Components}
\label{sec:PC}
We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
%
\begin{equation}
x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
\label{eq:mmutoxe}
\end{equation}
%
where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model. We obtain the PC basis functions $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements 
%
\beq
F_{ij} = \sum_l \frac{2 l+1}{2} \frac{\partial \ln C_l^{EE}}{\partial x_e(z_i)}\frac{\partial \ln C_l^{EE}}{\partial x_e(z_j)} = \sum_a \frac{ S_a(z_i) S_a(z_j)}{\sigma_a^2},
\eeq
%
where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the expected 
%
% added ``expected" to distinguish this from sigma(ma)
%
variances of the PCs.
The Fisher matrix is computed at a fiducial model which we take to be $x_e^{\rm fid}(z)=0.15$ in the parameterized range.
 Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the PCs to be computed with Planck 2015 rather than WMAP best-fit 5 remaining $\Lambda$CDM parameters, which results in minor differences between our $S_a$ and those of  Ref.~\cite{Heinrich:2016ojb}.

We then rank-order the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ in the fiducial analysis.\footnote{We also test the robustness of our fiducial analysis in \S \ref{sec:modelindependent} by  taking 
 for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.}

 In Fig.~\ref{fig:xe}
 %
 % this seems to be in the main text now
 %\footnote{There are very small differences between the PCs here and those shown in Ref.~\cite{Heinrich:2016ojb}, which can be attributed to a chance in the best-fit cosmology (WMAP to Planck 2015) when constructing the PCs.}
 %
 (top), we show the corresponding fiducial basis functions $S_a(z)$.  
Notice that the series resembles a Fourier decomposition of $x_e(z)$ rank ordered from low to high frequency.
This is because high frequency variations in $x_e$ leave little observable imprint in $C_l^{EE}$.  
As a consequence  the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
For the Planck data set, most of the information on the ionization history comes from contraints on the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}. 
For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a tanh transition width $\Delta z = 0.5$, given the helium fraction
 \beq
 f_{\rm He} = \frac{n_{\rm He}}{n_{H}} = \frac{m_{\rm H}}{m_{\rm He}} \frac{Y_p}{1 - Y_p}, 
 \eeq
as the ratio of the helium to hydrogen number density, where $Y_p$ is the helium mass fraction, chosen to be consistent with big bang nucleosynthesis for a given baryon density.

@@

 \begin{figure}
          \includegraphics[width=0.48\textwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
          \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$. We assume fully ionized Universe at $z < 6$. \textit{Bottom}: The cumulative optical depth from $z$ to $z_{\rm max}$ of a unit amplitude PC. } 
          \label{fig:xe}
\end{figure}

@@

\subsection{Effective PC Likelihood}
\label{sec:KDE}
The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data. In the following, we briefly recap the kernel density estimate (KDE) technique used to build this likelihood, as well as a Gaussian approximation, and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.

The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~(\ref{eq:xetommu}). Since $\mathbf{m}$ can take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm KDE}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
% 1.06 cut w/ physicality prior
% 1.072 no cut
For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
%Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors.

An even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
%
\begin{equation}
 {\cal L}_{\rm Gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{n_{\rm PC}} | \mathbf{C}| }}.
 \label{eq:gaussian}
 \end{equation}
%

For any set of model parameters ${\bf p}$ with their given prior probability distribution
$P({\bf p})$, their posterior distribution can be approximated as 
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p).
\end{equation}
with PC\,=\,KDE or Gaussian.
This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
In \S~\ref{sec:effective_likelihood} we test the effective KDE and Gaussian approaches against a direct implementation
for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

@@

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{align}
\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)}.
\label{eq:cumtau}
\end{align}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter. 
Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
\begin{align}
\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^5 m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
\label{eq:cumtauPC}
\end{align}
where $\tau_a$ and $\tau_{\rm fid}$ are defined by employing Eq.~({\ref{eq:mmutoxe}}) in
Eq.~(\ref{eq:cumtau}) using the fiducial model 
for $H(z)$ and number densities.
 When we omit the redshift arguments, we
implicitly mean the total range, e.g. 
 $
 \tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max}).
 $
 \ch{CH: what about recaling for different cosmological model?}

 In Fig.~\ref{fig:xe}, we display $\tau_a(z,z_{\rm max})$.
 Notice that positive values of the first component $m_1$ mainly provides the optical depth that is accumulated at high redshift whereas {\it negative} values of $m_2$ provides the same for low redshift.    Since these two are the best measured components, the CMB mainly determines the amount of low vs.~high redshift ionization.
The underlying reason is that in $C_l^{EE}$ the higher the redshift, the
larger the relative contribution at higher $l$ due to the size of the horizon when the photons scattered.  For the less well measured PCs with $a>2$, the contributions to the total optical depth rapidly diminish and represent finer distinctions in redshift for the cumulative optical depth and multipole for the underlying $C_l^{EE}$ spectrum.

In the usual approach to constraining reionization in $\Lambda$CDM, one places an implicit prior on the shape of the cumulative optical depth by choosing a ``tanh" or near step function reionization (see \S \ref{sec:example1}) and then extracts a single constraint on the total optical depth.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
unphysical ionization fractions where $0 \le x_e \le x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a  prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\label{eq:prior_sum}
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
Here we take $x_e^{\rm max} = 1+ f_{\rm He} $ to account for the contribution of singly ionized helium.
In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
% I think it can't depend on the data but does depend on the truncation
%the Planck 2018 data \wh{and
our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.  We study its impact and robustness in \S \ref{sec:modelindependent}.

%
Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for a physical ionization history.  
In other words, no physical model is excluded by these priors but some unphysical models are included.  In \S \ref{sec:note_on_priors} we further explore the role of these and other choices of PC priors, and in \S\ref{sec:effective_likelihood} we test PC constraints against exact constraints for two example models.

@@

\begin{table}[H]
\centering
\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$,  and correlation matrix $R_{ab}$. $\tau_a$ is the optical depth contribution for $m_a=1$ relative to 
$\tau_{\rm fid}=0.08626$.}

\label{tab:PC_stats}
\begin{tabular}{|r |r r r r r|}
\hline
